A meta-analysis of the impacts of best management practices on nonpoint source pollutant concentration.

### 2024-05-30 frontiers_submission/frontiers_submission.tex
prep for resubmission

Our systematic review identified a total of 33 studies and 125 effect sizes on FIB, 24 studies and 50 effect sizes on TN, 31 studies and 88 effect sizes for DIN, 31 studies and 61 effect sizes for TP, 17 studies and 36 effect sizes for PO\textsubscript{4}, and 33 studies with 125 effect sizes for TSS.
The majority of studies were identified as smaller scaled lot or field studies (Figure \ref{fig:bmpsummary} A).
FIB studies had a roughly equal proportion of large watershed/catchment studies and studies conducted at the community/farm scale or smaller.
We also identified that the majority of studies (all parameters) were conducted on urban or non-agricultural runoff (Figure \ref{fig:bmpsummary} B).
We did identify a wide variety of BMPs in the review, but it did not appear that any particular type of BMP was responsible for the majority of studies for any given parameter (Figure \ref{fig:bmpsummary} C).
Our review was restricted to studies published after between 1999 and 2023.
The number of studies published for each parameter were roughly uniformly distributed over time (Figure \ref{fig:bmptemporal} A) and are not indicative of increases or decreases in the number of published studies.
Study length was strongly skewed for all parameters (Figure \ref{fig:bmptemporal} B). Median study lengths were 3 (DIN), 2 (FIB), 2.5 (PO\textasciitilde4), 2.5 (TN), 2.5 (TP), and 2 (TSS) years.
There appears to be a strong clustering of BMP studies in the mid-Atlantic region (North Carolina, Virginia, Maryland) with other states sparsely represented or completely absent from the review (Figure \ref{fig:studymap}).

@@

There were only 19 studies and 63 FIB effect sizes available to model after removal of studies and effects that only reported BMP\textsubscript{eff}.
The overall mean effect (estimated with the intercept only multilevel random effects model) showed significant mean reductions in FIB (ROM = 0.85, 95\% CI {[}0.36, 1.34{]}; BMP\textsubscript{eff} = 57.4\%, 95\% CI {[}30.4\%, 73.9\%{]}; Figure \ref{fig:overalleffect}) resulting from BMPs.
Total heterogeneity was moderate with a relatively large amount of heterogeneity observed due to differences within studies (\textit{I\textsuperscript{2}\textsubscript{total}} = 53.54, \textit{I\textsuperscript{2}\textsubscript{study}} = 10.03, \textit{I\textsuperscript{2}\textsubscript{effect}} = 43.51).

AICc scores included log transformed influent concentration and the aridity index\texttimes BMP subcategory interaction as moderators for the FIB model (Table \ref{tab:modelaic}).
Moderator terms and interactions explained a high proportion of effect size variance (\textit{R\textsuperscript{2}\textsubscript{marginal}} =0.89) in the FIB model.
Increased influent concentrations (\(\beta\) = 0.25, 95\% CI {[}0.14, 0.37{]}) resulted in significantly larger ROM effect for FIB (Figure \ref{fig:fibeffect}).
Compared to the baseline aridity index\texttimes detention BMP subcategory interaction, infiltration (\(\beta\) = -29.90, 95\% CI {[}-50.34, -9.47{]}), livestock management (\(\beta\) = -30.37, 95\% CI {[}-50.93, -9.81{]}), and treatment (\(\beta\) = -30.33, 95\% CI {[}-49.62, -11.03{]}) interactions had significantly smaller slopes.
However, the data had uneven coverage of BMP subcategories across the aridity index.
Effects for detention BMPs were clustered in humid climates (aridity index \textgreater{} 0.65) and the resulting estimate for the baseline interaction (\(\beta\) = 32.63, 95\% CI {[}12.57, 52.69{]}) may not be reliable.

@@

We identified 13 eligible TN studies and 14 DIN studies and 31 and 44 effect sizes respectively that could be included in the regression model.
Overall effects showed that BMPs resulted in significant mean reductions in TN (ROM = 0.42, 95\% CI {[}0.21, 0.62{]}; BMP\textsubscript{eff} = 34.0\%, 95\% CI {[}18.7\%, 46.4\%{]}; Figure \ref{fig:overalleffect}) but not in DIN (ROM = 0.64, 95\% CI {[}-0.08, 1.35{]}; BMP\textsubscript{eff} = 47.1\%, 95\% CI {[}-8.1\%, 74.1\%{]}; Figure \ref{fig:overalleffect}).
Heterogeneity was high for TN with a large proportion of heterogeneity attributed to within study effect (\textit{I\textsuperscript{2}\textsubscript{total}} = 77.12, \textit{I\textsuperscript{2}\textsubscript{study}} = 23.2, \textit{I\textsuperscript{2}\textsubscript{effect}} = 53.92).
The DIN model had even higher heterogeneity with a larger proportion attributed between studies (\textit{I\textsuperscript{2}\textsubscript{total}} = 99.51, \textit{I\textsuperscript{2}\textsubscript{study}} = 83.53, \textit{I\textsuperscript{2}\textsubscript{effect}} = 15.97).
AICc scores indicated that none of the moderators resulted in substantial improvement over the intercept only model (Table \ref{tab:modelaic}).

@@

We found 17 TP studies with 37 effect sizes and 9 PO\textsubscript{4} studies with 21 effect sizes for inclusion in regression models.
There was a significant overall reduction found for TP (ROM = 0.40, 95\% CI {[}0.03, 0.76{]}; BMP\textsubscript{eff} = 32.7\%, 95\% CI {[}3.4\%, 53.2\%{]}) but no evidence of negative or positive effect for PO\textsubscript{4} (ROM = -0.18, 95\% CI {[}-0.56, 0.19{]}; BMP\textsubscript{eff} = -20.1\%, 95\% CI {[}-75.3\%, 17.7\%{]}).
For both the TP and PO\textsubscript{4} models, heterogenity was high, with moderate to high within study variance and low to moderate between study variance (TP: \textit{I\textsuperscript{2}\textsubscript{total}} = 96.13, \textit{I\textsuperscript{2}\textsubscript{study}} = 32.15, \textit{I\textsuperscript{2}\textsubscript{effect}} = 63.99; PO\textsubscript{4}: \textit{I\textsuperscript{2}\textsubscript{total}} = 97.28, \textit{I\textsuperscript{2}\textsubscript{study}} = 33.36, \textit{I\textsuperscript{2}\textsubscript{effect}} = 63.92).
The best model for both parameters only included influent as a moderator (Table \ref{tab:modelaic}).
Moderators explained a relatively small amount of variance for both models (TP: \textit{R\textsuperscript{2}\textsubscript{marginal}} =0.12, PO\textsubscript{4}: \textit{R\textsuperscript{2}\textsubscript{marginal}} =0.35).
Influent concentration (\(\beta\) = 0.23, 95\% CI {[}-0.035, 0.49{]}) was not significant at the 95\% confidence level for the TP model (Figure \ref{fig:phoseffect}; Table S4).
Influent concentration (\(\beta\) = 0.27, 95\% CI {[}0.085, 0.44{]}) was significant for the PO\textsubscript{4} model (Figure \ref{fig:phoseffect}; Table S5).

@@

Our systematic review revealed strong spatial disparities in published BMP studies (Figure \ref{fig:studymap}).
Similar spatial disparities have been identified and discussed in \citet{kochNitrogenRemovalStormwater2014}, \citet{grudzinskiDoesRiparianFencing2020} and can be problematic for extrapolating results to other regions of interest.
Inconsistent spatial coverage presents a challenge for disentangling confounding spatially correlated predictors such as climate and soil due to poor representation within the dataset.
\citet{horvathEffectsRegionalClimate2023} found overlapping BMP type and climate groups within their dataset that reduce the ability to distinguish effects due to either BMP type or climate.
Similarly, we found detention type BMPs clustered only in humid climates (high aridity index) which reduces our confidence in extrapolating the interaction between BMP types and aridity index for FIB (Figure \ref{fig:fibeffect}).
Not only were there spatially disparities, but we observed that the relative distribution of aridity index values does not resemble the distribution of aridity values across the U.S. (Figure \ref{fig:aiindex}).
Our review indicates that BMP studies are over represented in the generally humid regions of the country and underrepresented the more arid regions.
Study scale, runoff sources and BMP types appeared well distributed, with the caveat that we are not aware of the actual distribution of these values in BMPs deployed across the country.

@@

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_summary} \caption{Summary of (A) study scale, (B) dominant runoff source, and (C) BMPs identified in the systematic review.}\label{fig:bmpsummary}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_temporal_summary} \caption{Number of studies identified in the systematic review summarized by (A) publication date and (B) study length.}\label{fig:bmptemporal}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/study_map} \caption{Distribution of studies identified in the systematic review by state and parameter.}\label{fig:studymap}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/overall_effect} \caption{Estimated effect sizes and intervals from the intercept only multilevel random effects model. Individual points represent studies, with size scaled by sampling variance. The point estimate with uncertaintity bars indicate the estimated overall effect, 95\% confidence intervals, and 95\% prediction intervals. Here, \textit{k} indicates the number of overall effects with the number of unique studies in parenthesis.}\label{fig:overalleffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/bac_predicted} \caption{Predicted marignal effect of influent FIB and aridity index (conditioned on BMP subcategory). Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:fibeffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/phos_predicted} \caption{Predicted marignal effect of influent pollutant concentration on TP (Panel A) and PO\textsubscript{4} (Panel B) reductions. Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:phoseffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=0.5\linewidth,]{../figures/aridity_density} \caption{Comparison of the relative distributions (density) of aridity index values across the U.S. against aridity index values for studies in the systematic review.}\label{fig:aiindex}
\end{figure}

>>>>

Our systematic review identified a total of 33 studies and 125 effect sizes on FIB, 24 studies and 50 effect sizes on TN, 31 studies and 88 effect sizes for DIN, 31 studies and 61 effect sizes for TP, 17 studies and 36 effect sizes for PO\textsubscript{4}, and 33 studies with 125 effect sizes for TSS.
The majority of studies were identified as smaller scaled lot or field studies (Figure \ref{fig:fig1} A).
FIB studies had a roughly equal proportion of large watershed/catchment studies and studies conducted at the community/farm scale or smaller.
We also identified that the majority of studies (all parameters) were conducted on urban or non-agricultural runoff (Figure \ref{fig:fig1} B).
We did identify a wide variety of BMPs in the review, but it did not appear that any particular type of BMP was responsible for the majority of studies for any given parameter (Figure \ref{fig:fig1} C).
Our review was restricted to studies published after between 1999 and 2023.
The number of studies published for each parameter were roughly uniformly distributed over time (Figure \ref{fig:fig2} A) and are not indicative of increases or decreases in the number of published studies.
Study length was strongly skewed for all parameters (Figure \ref{fig:fig2} B). Median study lengths were 3 (DIN), 2 (FIB), 2.5 (PO\textasciitilde4), 2.5 (TN), 2.5 (TP), and 2 (TSS) years.
There appears to be a strong clustering of BMP studies in the mid-Atlantic region (North Carolina, Virginia, Maryland) with other states sparsely represented or completely absent from the review (Figure \ref{fig:fig3}).

@@

There were only 19 studies and 63 FIB effect sizes available to model after removal of studies and effects that only reported BMP\textsubscript{eff}.
The overall mean effect (estimated with the intercept only multilevel random effects model) showed significant mean reductions in FIB (ROM = 0.85, 95\% CI {[}0.36, 1.34{]}; BMP\textsubscript{eff} = 57.4\%, 95\% CI {[}30.4\%, 73.9\%{]}; Figure \ref{fig:fig4}) resulting from BMPs.
Total heterogeneity was moderate with a relatively large amount of heterogeneity observed due to differences within studies (\textit{I\textsuperscript{2}\textsubscript{total}} = 53.54, \textit{I\textsuperscript{2}\textsubscript{study}} = 10.03, \textit{I\textsuperscript{2}\textsubscript{effect}} = 43.51).

AICc scores included log transformed influent concentration and the aridity index\texttimes BMP subcategory interaction as moderators for the FIB model (Table \ref{tab:modelaic}).
Moderator terms and interactions explained a high proportion of effect size variance (\textit{R\textsuperscript{2}\textsubscript{marginal}} =0.89) in the FIB model.
Increased influent concentrations (\(\beta\) = 0.25, 95\% CI {[}0.14, 0.37{]}) resulted in significantly larger ROM effect for FIB (Figure \ref{fig:fig5}).
Compared to the baseline aridity index\texttimes detention BMP subcategory interaction, infiltration (\(\beta\) = -29.90, 95\% CI {[}-50.34, -9.47{]}), livestock management (\(\beta\) = -30.37, 95\% CI {[}-50.93, -9.81{]}), and treatment (\(\beta\) = -30.33, 95\% CI {[}-49.62, -11.03{]}) interactions had significantly smaller slopes.
However, the data had uneven coverage of BMP subcategories across the aridity index.
Effects for detention BMPs were clustered in humid climates (aridity index \textgreater{} 0.65) and the resulting estimate for the baseline interaction (\(\beta\) = 32.63, 95\% CI {[}12.57, 52.69{]}) may not be reliable.

@@

We identified 13 eligible TN studies and 14 DIN studies and 31 and 44 effect sizes respectively that could be included in the regression model.
Overall effects showed that BMPs resulted in significant mean reductions in TN (ROM = 0.42, 95\% CI {[}0.21, 0.62{]}; BMP\textsubscript{eff} = 34.0\%, 95\% CI {[}18.7\%, 46.4\%{]}; Figure \ref{fig:fig4}) but not in DIN (ROM = 0.64, 95\% CI {[}-0.08, 1.35{]}; BMP\textsubscript{eff} = 47.1\%, 95\% CI {[}-8.1\%, 74.1\%{]}; Figure \ref{fig:fig4}).
Heterogeneity was high for TN with a large proportion of heterogeneity attributed to within study effect (\textit{I\textsuperscript{2}\textsubscript{total}} = 77.12, \textit{I\textsuperscript{2}\textsubscript{study}} = 23.2, \textit{I\textsuperscript{2}\textsubscript{effect}} = 53.92).
The DIN model had even higher heterogeneity with a larger proportion attributed between studies (\textit{I\textsuperscript{2}\textsubscript{total}} = 99.51, \textit{I\textsuperscript{2}\textsubscript{study}} = 83.53, \textit{I\textsuperscript{2}\textsubscript{effect}} = 15.97).
AICc scores indicated that none of the moderators resulted in substantial improvement over the intercept only model (Table \ref{tab:modelaic}).

@@

We found 17 TP studies with 37 effect sizes and 9 PO\textsubscript{4} studies with 21 effect sizes for inclusion in regression models.
There was a significant overall reduction found for TP (ROM = 0.40, 95\% CI {[}0.03, 0.76{]}; BMP\textsubscript{eff} = 32.7\%, 95\% CI {[}3.4\%, 53.2\%{]}) but no evidence of negative or positive effect for PO\textsubscript{4} (ROM = -0.18, 95\% CI {[}-0.56, 0.19{]}; BMP\textsubscript{eff} = -20.1\%, 95\% CI {[}-75.3\%, 17.7\%{]}).
For both the TP and PO\textsubscript{4} models, heterogenity was high, with moderate to high within study variance and low to moderate between study variance (TP: \textit{I\textsuperscript{2}\textsubscript{total}} = 96.13, \textit{I\textsuperscript{2}\textsubscript{study}} = 32.15, \textit{I\textsuperscript{2}\textsubscript{effect}} = 63.99; PO\textsubscript{4}: \textit{I\textsuperscript{2}\textsubscript{total}} = 97.28, \textit{I\textsuperscript{2}\textsubscript{study}} = 33.36, \textit{I\textsuperscript{2}\textsubscript{effect}} = 63.92).
The best model for both parameters only included influent as a moderator (Table \ref{tab:modelaic}).
Moderators explained a relatively small amount of variance for both models (TP: \textit{R\textsuperscript{2}\textsubscript{marginal}} =0.12, PO\textsubscript{4}: \textit{R\textsuperscript{2}\textsubscript{marginal}} =0.35).
Influent concentration (\(\beta\) = 0.23, 95\% CI {[}-0.035, 0.49{]}) was not significant at the 95\% confidence level for the TP model (Figure \ref{fig:fig6}; Table S4).
Influent concentration (\(\beta\) = 0.27, 95\% CI {[}0.085, 0.44{]}) was significant for the PO\textsubscript{4} model (Figure \ref{fig:fig6}; Table S5).

@@

Our systematic review revealed strong spatial disparities in published BMP studies (Figure \ref{fig:fig3}).
Similar spatial disparities have been identified and discussed in \citet{kochNitrogenRemovalStormwater2014}, \citet{grudzinskiDoesRiparianFencing2020} and can be problematic for extrapolating results to other regions of interest.
Inconsistent spatial coverage presents a challenge for disentangling confounding spatially correlated predictors such as climate and soil due to poor representation within the dataset.
\citet{horvathEffectsRegionalClimate2023} found overlapping BMP type and climate groups within their dataset that reduce the ability to distinguish effects due to either BMP type or climate.
Similarly, we found detention type BMPs clustered only in humid climates (high aridity index) which reduces our confidence in extrapolating the interaction between BMP types and aridity index for FIB (Figure \ref{fig:fig5}).
Not only were there spatially disparities, but we observed that the relative distribution of aridity index values does not resemble the distribution of aridity values across the U.S. (Figure \ref{fig:fig7}).
Our review indicates that BMP studies are over represented in the generally humid regions of the country and underrepresented the more arid regions.
Study scale, runoff sources and BMP types appeared well distributed, with the caveat that we are not aware of the actual distribution of these values in BMPs deployed across the country.

@@

\begin{figure}
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig1} \caption{Summary of (A) study scale, (B) dominant runoff source, and (C) BMPs identified in the systematic review.}\label{fig:fig1}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig2} \caption{Number of studies identified in the systematic review summarized by (A) publication date and (B) study length.}\label{fig:fig2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig3} \caption{Distribution of studies identified in the systematic review by state and parameter.}\label{fig:fig3}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig4} \caption{Estimated effect sizes and intervals from the intercept only multilevel random effects model. Individual points represent studies, with size scaled by sampling variance. The point estimate with uncertaintity bars indicate the estimated overall effect, 95\% confidence intervals, and 95\% prediction intervals. Here, \textit{k} indicates the number of overall effects with the number of unique studies in parenthesis.}\label{fig:fig4}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig5} \caption{Predicted marignal effect of influent FIB and aridity index (conditioned on BMP subcategory). Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:fig5}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{frontiers_submission_files/figure-latex/fig6} \caption{Predicted marignal effect of influent pollutant concentration on TP (Panel A) and PO\textsubscript{4} (Panel B) reductions. Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:fig6}
\end{figure}

\begin{figure}[p]
\includegraphics[width=0.5\linewidth,]{frontiers_submission_files/figure-latex/fig7} \caption{Comparison of the relative distributions (density) of aridity index values across the U.S. against aridity index values for studies in the systematic review.}\label{fig:fig7}
\end{figure}
